---
title: "Lab 02 - Plastic waste"
author: "Julie Garrison"
date: "Jan 4 2025"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

### Exercise 1

Plot, using histograms, the distribution of plastic waste per capita faceted by continent. What can you say about how the continents compare to each other in terms of their plastic waste per capita?

Africa, Asia, South America: These continents mostly show low levels of plastic waste per capita, likely due to less industrialization and lower levels of single-use plastic consumption.
Europe: Europe shows a moderate distribution, likely due to higher consumption but also potentially better waste management practices.
North America: North America shows higher plastic waste per capita, likely due to high consumer behavior and many single-use plastics.

```{r plastic-waste-continent}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2, fill = "skyblue", color = "black") +
  facet_wrap(~ continent) +
  labs(
    title = "Distribution of Plastic Waste Per Capita by Continent",
    x = "Plastic Waste Per Capita",
    y = "Count"
  )

```

### Exercise 2
Recreate the density plots above using a different (lower) alpha level that works better for displaying the density curves for all continents.
```{r plastic-waste-density}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent
  )
) +
  geom_density(alpha = 0.2) +
 labs(
    title = "Density of Plastic Waste per Capita",
    subtitle = "by continent",
    x = "Plastic Waste per Capita",
    color = "Continent"
  )
```

Color and fill are mapped to continent because they are determined by the data. Alpha is set globally in geom_density() because it applies uniformly and does not depend on the data.

### Exercise 3

Violin plots are useful for understanding the detailed shape of the data, while box plots are good at summarizing key statistical values and highlighting outliers more clearly.


```{r plastic-waste-violin}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
  )
) +
  geom_violin()
```

### Exercise 4

Visualize the relationship between plastic waste per capita and mismanaged plastic waste per capita using a scatterplot. Describe the relationship.

4.1 
As plastic waste per capita increases, mismanaged plastic waste per capita tends to increase

```{r plastic-waste-mismanaged}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = mismanaged_plastic_waste_per_cap,
    y = plastic_waste_per_cap,
    color = continent,

  )
) +
  geom_point()
```

4.2
Developed regions (Europe and North America) tend to have higher plastic waste per capita with relatively lower mismanaged plastic waste, suggesting better management practices.
Developing regions (Africa and Asia) tend to have lower plastic waste per capita but proportionally higher levels of mismanaged plastic waste, highlighting infrastructure gaps.

4.3
Visualize the relationship between plastic waste per capita and total population as well as plastic waste per capita and coastal population. You will need to make two separate plots. Do either of these pairs of variables appear to be more strongly linearly associated?
```{r plastic-waste-mismanaged-by-pop}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = mismanaged_plastic_waste_per_cap,
    y = total_pop,
    color = continent,
  )
) +
  geom_point()

ggplot(
  data = plastic_waste,
  mapping = aes(
    x = mismanaged_plastic_waste_per_cap,
    y = coastal_pop,
    color = continent,
  )
) +
  geom_point()
```

The relationship between mismanaged plastic waste per capita and coastal population appears to be more linearly associated than the relationship with total population

### Exercise 5



```{r recreate-viz}
ggplot(
  data = plastic_waste, 
  mapping = aes(
    x = coastal_pop/total_pop, 
    y = plastic_waste_per_cap, 
    color = continent
    )) +
  geom_point(alpha = 0.7) +
  geom_smooth( color = "black", se = TRUE) +
  coord_cartesian(ylim = c(0, .8)) +

  

  labs(
    title = "Plastic waste vs. coastal population proportion",
    subtitle = "by continent",
    x = "Coastal population proportion (Coastal / total population)",
    y = "Plastic waste per capita",
    color = "Continent"
  )
```


